Statistical Analysis

### 2024-04-17 code/analysis-code/statistical-analysis.qmd
Part 5 Updates
Added figure captions for both statistical analysis sections. Converted tibbles to organized tables for the rendered manuscript. Added a legend to the bootstrapping graph.

# save fit results table  
table_file1 = here("results", "tables", "resulttable1.rds")
saveRDS(lmtable1, file = table_file1)
```
There appears to be a very strong relationship between total number of women using this birth control method and the outcome based on the coefficient estimate of 87.65998.

@@

# save fit results table  
table_file2 = here("results", "tables", "resulttable2.rds")
saveRDS(lmtable2, file = table_file2)

@@

# save fit results table  
table_file8 = here("results", "tables", "resulttable8.rds")
saveRDS(lmtable8, file = table_file8)
```
This model shows the strongest predictor is the number of women using short acting reversible contraception with a huge coefficient of 85.6433. The next strongest predictor is the European region followed by the Latin America/the Caribbean region and the Asia region, which is a shift from the model with the singular region predictor. The Oceania region now has a negative association with the outcome. Both of the marriage status predictors are associated with a decrease in the outcome. The coefficient estimates for the rate of maternal death due to unintended pregnancies and abortion both flipped from the single predictor models to show a positive association with the outcome.

@@

# create visualization with all 5 variables 
figure2 <- ggplot(finalbootstrap, aes(x = observed, y = point_estimates)) +
              geom_point(color = "black") +  
              geom_point(aes(y = medians), color = "coral") + 
              geom_errorbar(aes(ymin = lower_bound, ymax = upper_bound), width = 0.1, color = "blue") + 
              geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "black") +  # add 45 degree line
              xlim(0, 50) + # limit the x-axis values
              ylim(0, 50) + # limit the y-axis values
              labs(x = "Observed Values", y = "Predicted Values", title = "Uncertainty in Predictions with Final Model")
figure2

# save final figure
figure_file2 = here("results", "figures", "final-figures", "bootstrappred.png")
ggsave(filename = figure_file2, plot=figure2) 
```
This figure shows that 95% confidence intervals, shown as dark blue lines, for the predictions that do not fall on the expected 45 degree line do not intersect with the line either. The dark blue lines show that the confidence intervals are narrow, which increases the validity of the conclusion that the model is overfitting the data because there is not much uncertainty in the predictions.

@@

# save fit results table  
table_file16 = here("results", "tables", "resulttable16.rds")
saveRDS(lmtable11, file = table_file16)
```
As expected, the coefficients are much smaller than the first full model we created without the regularization penalty, but all 5 predictors remain in the model. The `sarc_standard` predictor still has the strongest relationship with the outcome. The percent of women currently married, rate of maternal deaths due to unintended pregnancies, and rate of maternal deaths to due to abortion all have weak negative associations with the outcome. The direction of the relationship between the rate of maternal deaths due to abortions and the outcome flipped from the first full regression model to this model, but the coefficient is so small that this difference is not significant.

@@

Once we decided that the full model with 5 predictors was the best choice, we used cross-validation with LASSO regression and random forest modeling to explore subset selection. The LASSO regression kept all 5 predictors, and the coefficients for the Asia, Europe, and Latin America/the Caribbean predictors were the largest behind the sarc_standard predictor. The smallest coefficient was assigned to the percent of currently married women predictor. The parameters were tuned for the LASSO regression, which increased our confidence in using a penalty of 0.1. The random forest model generated the same RMSE of 4.361363 as the LASSO regression model, but it has a higher R-squared value of 0.8369754. Since the random forest model explains 83.69% of the variation between the predicted and observed outcomes compared to 82.06% explained by the LASSO regression model, it is slightly better at predicting the outcome. After tuning the parameters for this model, the number of randomly selected predictors that produces the lowest RMSE is 4, not 5. While we cannot be sure which predictor the model is suggesting to remove, we hypothesize that it is the percent of currently married women based on the information from the LASSO regression. However, the improvement in RMSE when using 4 predictors instead of 5 is minimal, and the minimal node size variable is much more impactful on the RMSE of the model. Based on all of this information, we suggest using the full model with all 5 predictors with a LASSO regression because it accurately portrays the importance of each predictor while providing a penalty to predictors that do not meaningfully contribute to the model. While this model has a slightly lower R-squared value compared to the random forest model, the evidence of overfitting is so strong that this difference is negligible. Cross-validation of the model with all 5 predictors showed that predictions from testing data deviated from the expected relationship with larger variance than the predictions from the training data. Since the model did not perform well when used on test data, we conclude that there is overfitting in our model. Unfortunately, it is likely that any of these models are not optimal choices for predicting our outcome based on the poor performance of our model on testing data. While we suggest using the LASSO regression with all 5 predictors to predict our outcome because it is slightly easier to interpret, future researchers who are willing to sacrifice interpretation for predictive power should use the random forest model with 4 predictors based on the slightly higher predictive performance shown by the R-squared performance metric.

>>>>

# create a professional format of the table to add to the manuscript 
kable_lmtable1 <- kable(lmtable1)

# save fit results table  
table_file1 = here("results", "tables", "resulttable1.rds")
saveRDS(kable_lmtable1, file = table_file1)
```
There appears to be a very strong relationship between total number of women using this birth control method and the outcome based on the coefficient estimate of 87.65998.

@@

# create a professional format of the table to add to the manuscript 
kable_lmtable2 <- kable(lmtable2)

# save fit results table  
table_file2 = here("results", "tables", "resulttable2.rds")
saveRDS(kable_lmtable2, file = table_file2)

@@

# create a professional format of the table to add to the manuscript 
kable_lmtable8 <- kable(lmtable8)

# save fit results table  
table_file8 = here("results", "tables", "resulttable8.rds")
saveRDS(kable_lmtable8, file = table_file8)
```
This model shows the strongest predictor is the number of women using short acting reversible contraception with a huge coefficient of 85.6433. The next strongest predictor is the European region followed by the Latin America/the Caribbean region and the Asia region, which is a shift from the model with the singular region predictor. The Oceania region now has a negative association with the outcome. Both of the marriage status predictors are associated with a decrease in the outcome. The coefficient estimates for the rate of maternal death due to unintended pregnancies and abortion both flipped from the single predictor models to show a positive association with the outcome.

@@

# create visualization with all 5 variables 
figure2 <- ggplot(finalbootstrap, aes(x = observed, y = point_estimates)) +
              geom_point(aes(shape = "Point Estimates", color = "black")) +  
              geom_point(aes(y = medians, shape = "Medians",color = "coral")) + 
              geom_errorbar(aes(ymin = lower_bound, ymax = upper_bound), width = 0.1, color = "blue") + 
              geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "black") +  # add 45 degree line
              xlim(0, 50) + # limit the x-axis values
              ylim(0, 50) + # limit the y-axis values
              labs(x = "Observed Values", y = "Predicted Values", title = "Uncertainty in Predictions with Final Model") +
              scale_color_manual(values = c("black" = "black", "coral" = "coral", "blue" = "blue"),
                                 labels = c("Point Estimates", "Medians", "Error Bars"))
figure2

# save final figure
figure_file2 = here("results", "figures", "final-figures", "bootstrappred.png")
ggsave(filename = figure_file2, plot=figure2) 
```
This figure shows that 95% confidence intervals, shown as dark blue lines, for the predictions that do not fall on the expected 45 degree line do not intersect with the line either. The point estimates, represented as black triangles, have decent overlap with the medians, represented as coral circles, which is a sign of good agreement between predicted and observed values. The dark blue lines show that the confidence intervals are narrow, which increases the validity of the conclusion that the model is overfitting the data because there is not much uncertainty in the predictions. Unfortunately, the model is inaccurate, but it is inaccurate without much variation.

@@

# create a professional format of the table to add to the manuscript 
kable_lmtable11 <- kable(lmtable11)

# save fit results table  
table_file16 = here("results", "tables", "resulttable16.rds")
saveRDS(kable_lmtable11, file = table_file16)
```
As expected, the coefficients are much smaller than the first full model we created without the regularization penalty, but all 5 predictors remain in the model. The `sarc_standard` predictor still has the strongest relationship with the outcome. The percent of women currently married, rate of maternal deaths due to unintended pregnancies, and rate of maternal deaths to due to abortion all have weak negative associations with the outcome. The direction of the relationship between the rate of maternal deaths due to abortions and the outcome flipped from the first full regression model to this model, but the coefficient is so small that this difference is not significant.

@@

Once we decided that the full model with 5 predictors was the best choice, we used cross-validation with LASSO regression and random forest modeling to explore subset selection. The LASSO regression kept all 5 predictors, and the coefficients for the Asia, Europe, and Latin America/the Caribbean predictors were the largest behind the sarc_standard predictor. The smallest coefficient was assigned to the percent of currently married women predictor. The parameters were tuned for the LASSO regression, which increased our confidence in using a penalty of 0.1. The random forest model generated the same RMSE of 4.361363 as the LASSO regression model, but it has a higher R-squared value of 0.8369754. Since the random forest model explains 83.69% of the variation between the predicted and observed outcomes compared to 82.06% explained by the LASSO regression model, it is slightly better at predicting the outcome. After tuning the parameters for this model, the number of randomly selected predictors that produces the lowest RMSE is 4, not 5. While we cannot be sure which predictor the model is suggesting to remove, we hypothesize that it is the percent of currently married women based on the information from the LASSO regression. However, the improvement in RMSE when using 4 predictors instead of 5 is minimal, but it is interesting that the graph suggests using a minimal node size of 4 if the model contains all 5 predictors. Based on all of this information, we suggest using the full model with all 5 predictors with a LASSO regression because it accurately portrays the importance of each predictor while providing a penalty to predictors that do not meaningfully contribute to the model. While this model has a slightly lower R-squared value compared to the random forest model, the evidence of overfitting is so strong that this difference is negligible. Cross-validation of the model with all 5 predictors showed that predictions from testing data deviated from the expected relationship with larger variance than the predictions from the training data. Since the model did not perform well when used on test data, we conclude that there is overfitting in our model. Unfortunately, it is likely that any of these models are not optimal choices for predicting our outcome based on the poor performance of our model on testing data. While we suggest using the LASSO regression with all 5 predictors to predict our outcome because it is slightly easier to interpret, future researchers who are willing to sacrifice interpretation for predictive power should use the random forest model with 4 predictors based on the slightly higher predictive performance shown by the R-squared performance metric.
